Stochastic Variance Reduced Policy Gradient

### 2018-01-25 svrpg.tex
finished proof (minus comments); next: add minibatch to algorithm exposition

\section{Convergence Guarantees}
\begin{assumption}\label{ass:bounded_score}
For each trajectory $\tau$, value of $\vtheta$ and gradient component:
\[
	|\score{\vtheta}{\tau}| \leq G
\]
\end{assumption}

@@

\begin{restatable}[]{lemma}{L-smoothness}\label{lemma:lsmooth}
Under Assumption \ref{ass:bounded_score}, $J(\vtheta)$ is L-smooth with Lipschitz constant:
\[
	L = \frac{2RG^2}{1-\gamma}
\]
\end{restatable}

\begin{restatable}[Convergence of the SVRPG algorithm]{theorem}{convergence}\label{theo:convergence}
Under Assumptions \ref{ass:bounded_score} and \ref{ass:M2}, the parameter vector $\vtheta_A$ returned by Algorithm \ref{alg:svrpg} after $T=m\times S$ iterations has, for some positive constants $\psi,D, F$ and for proper choice of step sizes $\alpha_t$, the following property:
\begin{align*}
	&\EVV[]
	{\norm[]{\nabla J(\vtheta_A)}^2} 
		\leq
		\frac{J(\vtheta^*)-J(\vtheta_0)}{\psi T} +
		\frac{Z}{N}
		+\frac{\Omega}{b}
\end{align*}
\end{restatable}

@@

+dG^2R^2W

@@

			+\frac{1}{b}\sum_{i=0}^{b-1} \score{\vtheta_t^{s+1}}{\tau_i}R(\tau_i) 
			-\frac{1}{b}\sum_{i=0}^{b-1}

@@

			+\frac{1}{b}\sum_{i=0}^{b-1}\left( 

@@

		\frac{1}{b}\sum_{i=0}^{b-1}\left(

@@

			\frac{1}{b^2}\sum_{i=0}^{b-1}

@@

			+\frac{1}{b^2}\sum_{i=0}^{b-1}

@@

	+\frac{L^2}{b}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}\nonumber\\

@@

	&\qquad+\frac{L^2}{b}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}
	+G^2\frac{R^2}{(1-\gamma)^2}\frac{1}{b^2}\sum_{i=0}^{b-1}\Ets{(\omega(\tau_i)-1)^2} \\

@@

	&\qquad+\frac{L^2}{b}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}
	+G^2\frac{R^2}{(1-\gamma)^2b^2}\sum_{i=0}^{b-1}\Varts{\omega(\tau_i)} \nonumber\\

@@

	&\qquad+\frac{L^2}{b}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}
	+G^2\frac{R^2W}{(1-\gamma)^2b},

@@

%Lemma 1
\begin{restatable}[]{lemma}{auxone}\label{lemma:aux1}
Under Assumptions \ref{ass:bounded_score} ans \ref{ass:M2}, the expected squared norm of the true gradient $\gradJ{\vtheta_t^{s+1}}$, for appropriate choices of $\alpha_t\geq0$ and $\beta_t>0$, can be bounded as follows:
\[
	\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \leq
	\frac{R_{t+1}^{s+1} - R_t^{s+1}}{\Psi_t} + \frac{d_t\Tr\Varts{\gradApp{\tilde{\vtheta}^s}{1}}}{N\Psi_t}
	+\frac{f_tW}{b\Psi_t},
\]
	where
\begin{align*}
	&R_t^{s+1}\coloneqq \Ets{J(\vtheta_t^{s+1}) - c_t\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}, \\
	&c_{m-1} = 0, \\
	&c_t = c_{t+1}\left(\frac{3}{2}+\alpha_t\beta_t+\alpha_t^2\right) + \alpha_t\frac{L^2}{b}+\frac{\alpha_t^2L^3}{2}, \\
	&\Psi_t = \alpha_t\left(1-\frac{c_{t+1}}{\beta_t}-\frac{\alpha_tL}{2}-\alpha_tc_{t+1}\right), \\
	&d_t = \frac{\alpha_t}{2}\left(1+2c_{t+1}+\alpha_t^2L+2\alpha_t^2c_{t+1}\right), \\
	&f_t = \alpha_t^2\frac{G^2R^2(L+2c_{t+1})}{2(1-\gamma)^2}.
\end{align*}
In particular, the following constraints on $\alpha_t$ and $\beta_t$ are sufficient:
\begin{align*}
&0\leq\alpha_t < \frac{2(1-\nicefrac{c_{t+1}}{\beta_t})}{L+2c_{t+1}} \\
&\beta_t > c_{t+1}.
\end{align*}
\end{restatable}
\begin{proof}
	We have:
	\begin{align}
	\Ets{J(\vtheta_{t+1}^{s+1})} 
	&\geq \Ets{J(\vtheta_t^{s+1})+\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}} - \frac{L}{2}\norm[]{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}}^2} \\
	&= \Ets{J(\vtheta_t^{s+1})+\alpha_t\dotprod{\gradJ{\vtheta_t^{s+1}}}{\gradBlack{\vtheta_t^{s+1}}} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \\
	&\geq
	\Ets{J(\vtheta_t^{s+1})+\alpha_t\norm[]{\gradJ{\vtheta_t^{s+1}}}^2 - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \nonumber\\
	&\qquad-
	\frac{\alpha_t}{2N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} -\frac{\alpha_tL^2}{2}\Ets[t-1]{\norm[]{\vtheta_t^{s+1} - \tilde{\vtheta}^s}^2},
\end{align}
where (11) is from the L-smoothness of $J(\vtheta)$, (12) is from the SVRPG update, and (13) is from Lemma \ref{lemma:aux0}.

@@

	-\frac{f_t}{b}W,\nonumber

@@

where (17) is from (13), (18) is from (16), and (19) is from Lemma \ref{lemma:aux0}.

@@

\convergence*
\begin{proof}
We prove the theorem for the following values of the constants:
\begin{align*}
& \psi \coloneqq \min_t\{\Psi_t\}, \\
& Z \coloneqq \frac{\max_t\{d_t\}V}{\psi} \\
& \Omega \coloneqq \frac{\max_t\{f_t\}W}{\psi}.
\end{align*}
The values of all other constants and the constraints on the step size are provided in Lemma \ref{lemma:aux1}.

...
\end{proof}

>>>>

\section{Convergence Guarantees}
\begin{assumption}\label{ass:bounded_score}
For each trajectory $\tau$, value of $\vtheta$ and dimensions $i,j$ there are positive constants $G,F<\infty$ such that:
\begin{align*}
	&\left|\nabla_{\theta_i}\log\pi_{\vtheta}(\tau)\right| \leq G \\
	&\left|\frac{\partial^2}{\partial\theta_i\partial\theta_j}\log\pi_{\vtheta}(\tau)\right| \leq F
\end{align*}
\end{assumption}

@@

\begin{restatable}[]{lemma}{L-smoothness}\label{lemma:lsmooth}
Under Assumption \ref{ass:bounded_score}, $J(\vtheta)$ is L-smooth with Lipschitz constant:
\[
	L = \frac{R(G^2+F)}{1-\gamma}
\]
\end{restatable}

\begin{restatable}[Convergence of the SVRPG algorithm]{theorem}{convergence}\label{theo:convergence}
Under Assumptions \ref{ass:bounded_score} and \ref{ass:M2}, the parameter vector $\vtheta_A$ returned by Algorithm \ref{alg:svrpg} after $T=m\times S$ iterations has, for some positive constants $\psi,\zeta, \xi$ and for proper choice of the step size, the following property:
\begin{align*}
	&\EVV[]
	{\norm[]{\nabla J(\vtheta_A)}^2} 
		\leq
		\frac{J(\vtheta^*)-J(\vtheta_0)}{\psi T} +
		\frac{\zeta}{N}
		+\frac{\xi}{B}
\end{align*}
\end{restatable}

@@

+\frac{G^2R^2W}{(1-\gamma)^2B}

@@

			+\frac{1}{B}\sum_{i=0}^{B-1} \score{\vtheta_t^{s+1}}{\tau_i}R(\tau_i) 
			-\frac{1}{B}\sum_{i=0}^{B-1}

@@

			+\frac{1}{B}\sum_{i=0}^{B-1}\left( 

@@

		\frac{1}{B}\sum_{i=0}^{B-1}\left(

@@

			\frac{1}{B^2}\sum_{i=0}^{B-1}

@@

			+\frac{1}{B^2}\sum_{i=0}^{B-1}

@@

	+\frac{L^2}{B}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}\nonumber\\

@@

	&\qquad+\frac{L^2}{B}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}
	+G^2\frac{R^2}{(1-\gamma)^2}\frac{1}{B^2}\sum_{i=0}^{B-1}\Ets{(\omega(\tau_i)-1)^2} \\

@@

	&\qquad+\frac{L^2}{B}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}
	+G^2\frac{R^2}{(1-\gamma)^2b^2}\sum_{i=0}^{B-1}\Varts{\omega(\tau_i)} \nonumber\\

@@

	&\qquad+\frac{L^2}{B}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}
	+G^2\frac{R^2W}{(1-\gamma)^2B},

@@

%Lemma 1
\begin{restatable}[]{lemma}{auxone}\label{lemma:aux1}
Under Assumptions \ref{ass:bounded_score} ans \ref{ass:M2}, the expected squared norm of the true gradient $\gradJ{\vtheta_t^{s+1}}$, for appropriate choices of $\alpha_t\geq0$ and $\beta_t>0$, can be bounded as follows:
\[
	\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \leq
	\frac{R_{t+1}^{s+1} - R_t^{s+1}}{\Psi_t} + \frac{d_tV}{N\Psi_t}
	+\frac{f_tW}{b\Psi_t},
\]
	where
\begin{align*}
	&R_t^{s+1}\coloneqq \Ets{J(\vtheta_t^{s+1}) - c_t\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}, \\
	&c_{m-1} = 0, \\
	&c_t = c_{t+1}\left(\frac{3}{2}+\alpha_t\beta_t+\alpha_t^2\right) + \alpha_t\frac{L^2}{b}+\frac{\alpha_t^2L^3}{2}, \\
	&\Psi_t = \alpha_t\left(1-\frac{c_{t+1}}{\beta_t}-\frac{\alpha_tL}{2}-\alpha_tc_{t+1}\right), \\
	&d_t = \frac{\alpha_t}{2}\left(1+2c_{t+1}+\alpha_t^2L+2\alpha_t^2c_{t+1}\right), \\
	&f_t = \alpha_t^2\frac{G^2R^2(L+2c_{t+1})}{2(1-\gamma)^2}.
\end{align*}
In particular, the following constraints on $\alpha_t$ and $\beta_t$ are sufficient:
\begin{align*}
&0\leq\alpha_t < \frac{2(1-\nicefrac{c_{t+1}}{\beta_t})}{L+2c_{t+1}} \\
&\beta_t > c_{t+1}.
\end{align*}
\end{restatable}
\begin{proof}
	We have:
	\begin{align}
	\Ets{J(\vtheta_{t+1}^{s+1})} 
	&\geq \Ets{J(\vtheta_t^{s+1})+\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}} - \frac{L}{2}\norm[]{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}}^2} \\
	&= \Ets{J(\vtheta_t^{s+1})+\alpha_t\dotprod{\gradJ{\vtheta_t^{s+1}}}{\gradBlack{\vtheta_t^{s+1}}} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \\
	&\geq
	\Ets{J(\vtheta_t^{s+1})+\alpha_t\norm[]{\gradJ{\vtheta_t^{s+1}}}^2 - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \nonumber\\
	&\qquad-
	\frac{\alpha_t}{2N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} -\frac{\alpha_tL^2}{2}\Ets[t-1]{\norm[]{\vtheta_t^{s+1} - \tilde{\vtheta}^s}^2},
\end{align}
where (11) is from the L-smoothness of $J(\vtheta)$, (12) is from the SVRPG update, and (13) is from Lemma \ref{lemma:aux0}.

@@

	-\frac{f_t}{b}W,\nonumber\\
	&\geq R_t^{s+1}
	-\frac{d_t}{N}V
	-\frac{f_t}{b}W,

@@

where (17) is from (13), (18) is from (16), and (19) is from Lemma \ref{lemma:aux0} and (20) is from Assumption \ref{ass:REINFORCE}.

@@

\convergence*
\begin{proof}
We prove the theorem for the following values of the constants:
\begin{align*}
& \psi \coloneqq \min_t\{\Psi_t\}, \\
& \zeta \coloneqq \frac{\max_t\{d_t\}V}{\psi} \\
& \xi \coloneqq \frac{\max_t\{f_t\}W}{\psi}.
\end{align*}
The values of all other constants and the constraints on the step size are provided in Lemma \ref{lemma:aux1}.

Starting from Lemma \ref{lemma:aux1}, summing over iterations of an epoch $s$ we get:
\begin{align}
\sum_{t=0}^{m-1}\Ets{\norm[]{\gradJ{(\vtheta_t^{s+1})}^2}}&\leq
 \frac{\sum_{t=0}^{m-1}\left(R_{t+1}^{s+1} - R_t^{s+1}\right)}{\psi} + \frac{m\zeta}{N} + \frac{m\xi}{B} \\
 %
 &= \frac{\sum_{t=0}^{m-1}\left(\Ets[t+1]{J(\vtheta_{t+1}^{s+1}) - c_{t+1}\norm[]{\vtheta_{t+1}^{s+1}-\tilde{\vtheta}^s}^2} - \Ets[t]{J(\vtheta_{t}^{s+1}) - c_{t}\norm[]{\vtheta_{t}^{s+1}-\tilde{\vtheta}^s}^2}\right)}{\psi} \nonumber\\
 &\qquad+ \frac{m\zeta}{N} + \frac{m\xi}{B} \\
 %
 &= \frac{\Ets[m]{\sum_{t=0}^{m-1}\left(J(\vtheta_{t+1}^{s+1}) - c_{t+1}\norm[]{\vtheta_{t+1}^{s+1}-\tilde{\vtheta}^s}^2 - J(\vtheta_{t}^{s+1}) - c_{t}\norm[]{\vtheta_{t}^{s+1}-\tilde{\vtheta}^s}^2\right)}}{\psi} \nonumber\\
 &\qquad+ \frac{m\zeta}{N} + \frac{m\xi}{B} \\
 %
 &= \frac{\Ets[m]{J(\vtheta_{m}^{s+1}) - c_{t+1}\norm[]{\vtheta_{m}^{s+1}-\tilde{\vtheta}^s}^2 - J(\vtheta_{0}^{s+1}) - c_{t}\norm[]{\vtheta_{0}^{s+1}-\tilde{\vtheta}^s}^2}}{\psi} \nonumber\\
 &\qquad+ \frac{m\zeta}{N} + \frac{m\xi}{B} \\
 %
 &= \frac{\Ets[m]{J(\tilde{\vtheta}^{s+1}) - J(\tilde{\vtheta}^{s})}}{\psi} + \frac{m\zeta}{N} + \frac{m\xi}{B}
 \end{align}
Next, summing over epochs:
\begin{align}
\sum_{s=0}^{S-1}\sum_{t=0}^{m-1}\Ets{\norm[]{\gradJ{(\vtheta_t^{s+1})}^2}}&\leq
\frac{\sum_{s=0}^{S-1}\Ets[m]{J(\tilde{\vtheta}^{s+1}) - J(\tilde{\vtheta}^{s})}}{\psi} + \frac{T\zeta}{N} + \frac{T\xi}{B} \\
%
&\leq
\frac{\EVV[]{J(\tilde{\vtheta}^{S}) - J(\tilde{\vtheta}^{0})}}{\psi} + \frac{T\zeta}{N} + \frac{T\xi}{B}
 \\
 %
 &\leq
 \frac{\EVV[]{J(\vtheta_{out})} - J(\tilde{\vtheta}^{0})}{\psi} + \frac{T\zeta}{N} + \frac{T\xi}{B} \\
 %
 &\leq
 \frac{J(\vtheta^*) - J(\vtheta^0)}{\psi} + \frac{T\zeta}{N} + \frac{T\xi}{B}
\end{align}
Finally, we consider the expectation \wrt all sources of randomness, including the uniform sampling of the output parameter:
\begin{align}
\EVV[]{\norm[]{\gradJ{(\vtheta_t^{s+1})}}^2} 
&=\frac{1}{T}\sum_{s=0}^{S-1}\sum_{t=0}^{m-1}\Ets{\norm[]{\gradJ{(\vtheta_t^{s+1})}}^2} \\
&\leq
\frac{J(\vtheta^*) - J(\vtheta^0)}{\psi T} + \frac{\zeta}{N} + \frac{\xi}{B}
\end{align}
\end{proof}
